```{r 17_setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE, fig.align="center")
```

# Applied Mediation Analysis

## Pre-class work {-}

Reading

- [How racial discrimination in law enforcement actually works](https://www.vox.com/2014/12/1/7311417/race-law-controls)


<br><br><br><br>


## Learning Goals {-}

- MEDI1: Formulate research questions that can be answered via mediation analysis
- MEDI2: Conduct and interpret results from a mediation analysis

<center>
<img src="images/course_flow6.svg" width="60%" height="60%">
</center>


<br><br><br><br>


## Assignment {-}

**A template RMarkdown document that you can start from is available [here](template_rmds/17-applied-mediation.Rmd).**

### Pre-class reading {-}

Write a short response to the article, [How racial discrimination in law enforcement actually works](https://www.vox.com/2014/12/1/7311417/race-law-controls). Use quotes from the article to discuss connections to mediation ideas and other concepts from our course. (Limit 500 words.)

Thoughtful responses will earn 2 M grades to apply to any objective of your choosing. (You can't use this to skip parts of Homework 5 or 6 though.)


<br><br>


### Applied mediation analysis {-}

The New York Times recently [featured](https://www.nytimes.com/2020/09/29/business/economy/pandemic-women-tenure.html) research by a group who studied the effect of tenure-clock stopping policies on promotion and job security-related outcomes.

We'll be looking at data from the study [Equal but Inequitable: Who Benefits from Gender-Neutral Tenure Clock Stopping Policies?](https://www.aeaweb.org/articles?id=10.1257/aer.20160613) by Antecol, Bedard, and Stearns.

You'll need to install the `mediation` package before getting started.

```{r}
install.packages("mediation")
```

The data and data information are available [on Moodle](https://moodle.macalester.edu/mod/folder/view.php?id=36923) (`aer_primarysample.csv` and `ReadMe.pdf`).

```{r}
library(dplyr)
library(ggplot2)
library(splines)
library(mediation)

tenure <- readr::read_csv("aer_primarysample.csv")
```

<br>


> **Context:** Tenure clock-stopping policies broadly relate to delaying the timeline for tenure review due to life events. Often this is related to becoming a parent. The COVID pandemic has also generated discussions about clock-stopping policies for both its standalone impacts on faculty and its intersectional effects with other factors like parenthood. Clock-stopping policies are intended to affect scholarly output which in turn affects ultimate awarding of tenure.

> **Broad question:** How do gender-neutral clock stopping policies affect the ultimate awarding of tenure, and how influential are publications in top journals in this process?

Information on variables starts at the bottom of Page 1 of `ReadMe.pdf` ("The independent variables of interest are..."). We will focus on the following variables:

- `gncs`: a gender-neutral clock-stopping policy is in place (treatment)
- `top_pubs5`: cumulative number of peer-reviewed publications in top-5 journals by year 5 since PhD completion (mediator)
- `tenure_policy_school`: indicator for getting tenure at the university with the clock-stopping policy


**Your task:**

- The authors assume that the variables `female + phd_rank + post_doc + ug_students + grad_students + faculty + full_av_salary + assist_av_salary + revenue + female_ratio + full_ratio` are enough to meet the exchangeability conditions discussed yesterday. Do you agree? Explain. State your final conditional exchangeability set.

- Use visualizations to build appropriate models for the mediator and for the outcome. Focus on correctly capturing nonlinearity rather than interactions. Call the mediator model `med_mod` and the outcome model `out_mod`.

- Use the `mediate()` function as demonstrated below to estimate the natural direct and indirect effects. (In the output, `ACME` stands for the indirect effect and `ADE` stands for the direct effect.)

```{r}
set.seed(451)
mediate_out <- mediate(med_mod, out_mod, treat = "gncs", mediator = "top_pubs5", robustSE = TRUE, sims = 1000)

summary(mediate_out)
```

- Decide whether NIE(0) or NIE(1) is more of interest to you and explain in a few sentences.

- Interpret the chosen NIE, the corresponding NDE, the total effect, and the proportion mediated quantity contextually.
