```{r appendix_causal_disc, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=TRUE)
```

# Causal Discovery with `pcalg` {-}

The [`pcalg` package](https://cran.r-project.org/package=pcalg) (pronounced "PC alg", short for "PC algorithm") allows us to perform casual discovery in R. The cornerstone causal discovery algorithm that we learned was the SGS algorithm. The PC algorithm (also named after its authors) uses the same ideas as SGS but performs the conditional independence tests in a better order for faster computation.

```{r eval=TRUE}
library(dplyr)
library(pcalg)

tenure <- readr::read_csv("../../data/aer_primarysample.csv")

tenure <- tenure %>%
    select(gncs, top_pubs5, tenure_policy_school, female, phd_rank, post_doc, ug_students, grad_students, faculty, full_av_salary, assist_av_salary, revenue, female_ratio, full_ratio)
```

```{r eval=TRUE}
suff_stat <- list(C = cor(tenure), n = nrow(tenure))
pc_tenure <- pc(suff_stat, indepTest = gaussCItest, labels = colnames(tenure), alpha = 0.01, skel.method = "stable.fast")

plot(pc_tenure, main = "")
```
