```{r hw5_setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

# Homework 5 {-}

<center>
**Due Sunday, October 4 at midnight CST on [Moodle](https://moodle.macalester.edu/mod/assign/view.php?id=34829)**

**Please turn in two PDF documents from class exercises.**

(If you have LaTeX installed, you can knit directly to PDF (preferred method). Otherwise, you can knit to HTML and "Print" the page to save it as a PDF.)
</center>


<br>


## Deliverables {-}

Please turn in exercises from the following class days:

- Topic 14: Sensitivity Analyses for Unmeasured Variables
- Topic 15: Causal Discovery

This should be a total of two PDF documents.

<br>


## Primary objectives {-}

The following learning objectives are the primary focus of this assignment:

- **`SENS1`**: Evaluate the sensitivity of findings to data quality and propose appropriate sensitivity analyses for a research investigation
- **`SENS2`**: Conduct and communicate the results of a sensitivity analysis for unmeasured confounding.
- **`DISC1`**: Manually conduct causal discovery using regression models.
- **`DISC2`**: Use output from causal discovery to enhance a causal analysis.


<br>


## Revisiting previous objectives {-}

You can revisit any of the following objectives (excludes `REGR1`, `IPTW1`, `TVTR1`, and `TVTR2`).

- **`EXCH1`:** Apply the concepts of marginal and conditional exchangeability to answer questions about (hypothetical) data on potential outcomes.
- **`EXCH2`:** Give examples of when marginal and conditional exchangeability would and would not hold in various data contexts.
- **`EXCH3`:** Explain why a direct comparison of the outcomes in the treated and untreated is misleading as an estimate of a causal effect.
- **`DESI1`:** Explain how randomized experiments relate to exchangeability.
- **`DESI2`:** Explain how quasi-experimental and general observational studies relate to exchangeability.
- **`DESI3`:** Compare the strengths and weaknesses of different study designs for answering a research question.
- **`PGRA1`:** Apply the Causal Markov assumption to express the joint distribution of data.
- **`CNCP1`:** Explain how causal and noncausal paths relate to exchangeability and causal effects.
- **`DSEP1`:** Apply d-separation to block noncausal paths in causal DAGs with and without unobserved variables.
- **`DSEP2`:** Apply strategies to deal with exchangeability problems caused by unobserved variables.
- **`DSEP3`:** Simulate data from a causal DAG under linear and logistic regression SEMs to check d-separation properties through regression modeling and visualization.
- **`DSEP4`:** Explain how d-separation relates to conditional exchangeability.
- **`TVTR1`**: Formulate research questions that can be answered in a time-varying treatment setting.
- **`TVTR2`**: Explain why regression does not generally work in time-varying settings with treatment-confounder feedback using d-separation ideas.

<br>

If you revisit any of these objectives, you must do so using the graph below. If revisiting an objective requires context for exactly what these variables represent, feel free to pick a context that is suitable for you.

```{r hw5_dag, eval=TRUE}
library(dagitty)
dag <- dagitty("dag {
bb=\"0,0,1,1\"
A [exposure,pos=\"0.400,0.350\"]
L [pos=\"0.200,0.350\"]
U1 [latent,pos=\"0.100,0.200\"]
U2 [latent,pos=\"0.100,0.500\"]
Y [outcome,pos=\"0.600,0.350\"]
A -> Y
L -> A
U1 -> L
U1 -> Y
U2 -> A
U2 -> L
}
")

plot(dag)
```

