```{r r_misc, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

# (PART) Miscellaneous {-}

# Additional R resources {-}

[DAGitty](http://www.dagitty.net/dags.html) does not display noncausal paths, which can often be helpful for trying to *approximately* block noncausal paths in the presence of unmeasured variables.

The code below shows a way to do this. For now, you have to define the DAG using a list of edges. (The example below shows the classic confounder triangle.)

If you're feeling adventurous, learn about the [`dagitty`](https://cran.r-project.org/package=dagitty) R package to see if you can write a function that takes the "Model code" from the DAGitty web interface (right side of the window) to directly create the `igraph` object.

```{r}
library(igraph)

# Create the edge list
el <- rbind(
    c("C", "A"),
    c("A", "Y"),
    c("C", "Y")
)
# Create the graph
dag <- graph_from_edgelist(el, directed = TRUE)

# Helper functions for displaying paths
display_paths <- function(g, paths) {
    adj <- as_adjacency_matrix(g)
    for (i in seq_along(paths)) {
        path <- paths[[i]]
        cat("Path ", i, ":   ", sep = "")
        draw_edge(adj_mat = adj, path = path)
    }
}

draw_edge <- function(adj_mat, path) {
    path_ids <- as_ids(path)
    arrow_vec <- rep("", length(path_ids))
    for (i in seq_len(length(path_ids)-1)) {
        node1 <- path_ids[i]
        node2 <- path_ids[i+1]
        if (adj_mat[node1,node2]==1) {
            ## Arrow is 1 -> 2
            arrow_vec[i] <- "->"
        } else {
            ## Arrow is 1 <- 2
            arrow_vec[i] <- "<-"
        }
    }
    path_string <- paste(path_ids, arrow_vec)
    path_string <- paste(path_string, collapse = " ")
    cat(path_string, "\n")
}

display_paths(dag, all_simple_paths(dag, from = "A", to = "Y", mode = "all"))
```
