---
title: "Review and Motivation"
author: "Your Name"
output: 
  html_document:
    toc: true
    toc_float: true
---

# Setup

```{r}
# Read in the data
# If necessary, install the "readr" package with: install.packages("readr")
chemo_study_data <- readr::read_csv("https://www.dropbox.com/s/vl06j75a8afw8ct/chemo_study.csv?dl=1")

# Load the ggplot2 package
```

# Exercise 1: Plots and confounders

The first step in any data analysis is to visualize your data. Let's refamiliarize ourselves with the `ggplot2` package in R. It may be helpful to have [this ggplot2 cheat sheet](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) open. Make sure to load the `ggplot2` package by including `library(ggplot2)` at the top of your RMarkdown document.

## Part a

Get a feel for the data by visualizing the distributions of all 4 variables individually (4 plots total). Your code should look something like below.

```{r}
ggplot(chemo_study_data, aes(x = pre_tumor_size)) + ???
```

## Part b

Is pre-treatment tumor size predictive of whether or not a patient received chemotherapy? Make a plot to assess this, and briefly state what conclusions can be drawn from the plot. (Hint: it will be helpful to look at the second page of the cheat sheet in the section labeled "Position Adjustments".)

```{r}
ggplot(chemo_study_data, aes(x = pre_tumor_size, fill = treated)) + ???
```

## Part c

Is pre-treatment tumor size predictive of whether or not a patient recovered? Make a plot, and briefly state your conclusions.

## Part d

A variable is a **confounder** if it is a common cause of both the treatment and outcome. This is shown in the diagram below. Given your results from parts b and c, could pre-treatment tumor size be a confounder of the relationship between chemotherapy treatment and recovery?

## Part e

Could *post*-treatment tumor size be a confounder of the relationship between ## Part chemotherapy treatment and recovery? Sup

rt your answer with plots and an explanation.


Note: the causal relationships of interest so far can be depicted in a **causal diagram**, shown below. An arrow between two variables indicates that one is a cause of the other (an arrow points from a cause to its effect).

```{r 01-fig-dag-chemo, fig.width=6, fig.height=2, fig.align="center", echo=FALSE, eval=TRUE}
par(mar = rep(0,4))
plot(1, type = "n", xaxt = "n", yaxt = "n", bty = "n", xlab = "", ylab = "", xlim = c(0,6), ylim = c(1,4))
text(c("Chemotherapy", "Pre-treatment tumor size", "Recovery"), x = c(1,3,5), y = c(1,4,1), cex = 1.1)
arrows(x0 = c(3,3), y0 = c(4,4)-0.2, x1 = c(1,5), y1 = c(1,1)+0.2, angle = 25, lwd = 4)
arrows(x0 = 1.7, y0 = 1, x1 = 4.5, y1 = 1, angle = 25, lwd = 4)
```


<br><br>


# Exercise 2: Implications of confounding

## Part a

What is the relationship between confounders and the saying "Correlation does not imply causation"?

## Part b

Explain how regression models can be used to remove the influence of confounding variables.

## Part c

Suppose that chemotherapy generally works to promote recovery. Suppose also that sicker patients tend to receive chemotherapy, and healthier patients tend to not receive chemotherapy. What do you expect 


<br><br>


# Exercise 3: Logistic regression models

We can model recovery using a **logistic regression model** (used when the outcome variable is binary). In R, we can fit a logistic regression model using code like the following:

```{r}
# Fit the model and store it in the "mod" object
mod <- glm(outcome_variable ~ predictor_variable1+predictor_variable2,
    family = "binomial", data = your_data)

# Display model output
summary(mod)

# Obtain confidence intervals (CIs) for model coefficients
# (95% CIs by default)
confint(mod)
```

## Part a

Fit a logistic regression model with only treatment as a predictor. Interpret the treatment coefficient. Is the sign of the coefficient what you expected? Why is this model misleading?

## Part b

Fit a logistic regression model that adjusts for pre-treatment tumor size. Interpret the treatment coefficient. Is the sign of the coefficient what you expected? Also report the 95% confidence interval for the treatment coefficient and what information you learn from it.

## Part c

Fit a logistic regression model that adjusts for pre- and post-treatment tumor size. Interpret the treatment coefficient. Is the sign of the coefficient what you expected?


(Side note: [This article](https://medium.com/@EpiEllie/having-confidence-in-confidence-intervals-8f881712d837) by Eleanor Murray has a great explanation of the most common misinterpretation of confidence intervals.)


<br><br>


# Exercise 4: Warming up to causal ideas

## Part a

Consider the statement: "The causal effect of chemotherapy on recovery rates (as compared to no chemotherapy) is a change (difference) of 10%." What do you think "causal effect" means? What would you like it to mean?

## Part b

Which model do you think best estimates the causal effect of chemotherapy on recovery rates? Explain.

