---
title: "Applied Analysis: Inverse Probability Weighting"
author: "Your Name"
output: 
  html_document:
    toc: true
    toc_float: true
---

# Setup

We'll be looking at the dataset explored in our WHATIF book to answer the question:

> **How does smoking cessation affect weight gain?**

The data are available in the `cidata` package. Install this package (you may need to install the `remotes` package first):

```{r}
remotes::install_github("malcolmbarrett/cidata") 
```

Our data come from the National Health and Nutrition Examination Survey Data I Epidemiologic Follow-up Study (NHEFS).

We will use the `nhefs_complete` dataset which has removed cases with missing data for some key variables. More information is available via `?nhefs_complete`. The variable codebook is available as `nhefs_codebook`.

```{r}
library(survey) # You may need to install the survey package
library(dplyr)
library(ggplot2)
library(cidata)

data(nhefs_complete) # Load the data
View(nhefs_codebook) # Look at the variable codebook
```

<br>

# Part 1: Causal graph analysis

Use the causal graph below as a starting point for the analysis. The selection node `S` indicates those for whom we have complete data. The treatment variable is `qsmk`, and the outcome variable is `wt82_71`.

- If this is the true causal graph underlying the data, can we achieve conditional exchangeability? Show your work fully.

```{r 12_dag, echo=FALSE, eval=TRUE}
library(dagitty)
dag <- dagitty("dag {
bb=\"0,0,1,1\"
\"age: age in 1971\" [pos=\"0.200,0.223\"]
\"qsmk: Quit smoking in 1971-82\" [exposure,pos=\"0.251,0.451\"]
\"smokeintensity: Cigs/day in 1971\" [pos=\"0.440,0.199\"]
\"smokeyrs: Years smoking in 1971\" [pos=\"0.733,0.184\"]
\"wt82_71: Weight change (kg) 1971-82\" [outcome,pos=\"0.763,0.448\"]
S [adjusted,pos=\"0.576,0.089\"]
\"age: age in 1971\" -> \"qsmk: Quit smoking in 1971-82\"
\"age: age in 1971\" -> \"wt82_71: Weight change (kg) 1971-82\"
\"age: age in 1971\" -> S
\"qsmk: Quit smoking in 1971-82\" -> \"wt82_71: Weight change (kg) 1971-82\"
\"smokeintensity: Cigs/day in 1971\" -> \"qsmk: Quit smoking in 1971-82\"
\"smokeintensity: Cigs/day in 1971\" -> \"wt82_71: Weight change (kg) 1971-82\"
\"smokeintensity: Cigs/day in 1971\" -> S
\"smokeyrs: Years smoking in 1971\" -> \"qsmk: Quit smoking in 1971-82\"
\"smokeyrs: Years smoking in 1971\" -> \"wt82_71: Weight change (kg) 1971-82\"
\"smokeyrs: Years smoking in 1971\" -> S
}
")
plot(dag)
```


<br><br>


# Part 2: Propensity score modeling

- Based on your causal graph analysis, fit an appropriate propensity score model. Use visualizations to inform the construction of your model. Accompany each visualization with a brief description of what it helps you decide.

- Use your model to compute appropriate weights, and add these weights to the dataset.

- Use "before and after" visualizations to compare balance of key variables before and after weighting. Write a few sentences summarizing your conclusions from this stage.

<br>

The code below adds logistic regression smoothing lines to `ggplot2` figures:

```{r}
ggplot(data, aes(x = X, y = y)) +
    geom_point() +
    geom_smooth(se = FALSE, color = "blue") +
    geom_smooth(formula = y~ns(x,2), method = "glm",
        method.args = list(family="binomial"),
        se = FALSE, color = "red"
    )
```

<br>

The (incomplete) code below is useful for computing appropriate weights:

```{r}
your_data <- your_data %>%
    mutate(
        ps = predict(ps_mod, newdata = your_data, type = "response"),
        ip_weight = case_when(
            A==1 ~ ???,
            A==0 ~ ???
        )
    )
```

<br>

You can incorporate weights into most `ggplot2` figures by adding `weight` to the `aes`thetics:

```{r}
aes(..., weight = ip_weight)
```


<br><br>


# Part 3: Modeling with IP weights

Assuming that you have added the `ip_weight` variable to the `nhefs_complete` dataset, the code below fits the model

$$
E[Y^a] = \beta_0 + \beta_1a
$$

(As discussed in 12.4 of WHATIF, this model is called a **marginal structural model (MSM)**.)

```{r}
# Set up information about weights
design <- svydesign(ids = ~0, weights = nhefs_complete$ip_weight, data = nhefs_complete)

msm_fit <- svyglm(
    wt82_71 ~ qsmk,
    data = data,
    design = design
)
summary(msm_fit)
confint(msm_fit)
```

- Fit a marginal structural model to estimate the overall average causal effect.

- Construct a visualization to explore if the causal effect of smoking cessation might differ by prior smoking intensity (`smokeintensity`). Briefly interpret your visualization.

- Fit a model to estimate how the average causal effect differs across smoking intensity. (Adapt the `svyglm()` formula in the same way you would for `lm()`.)

- Using both the confidence intervals and effect magnitudes, discuss the results of your analysis in a contextually meaningful way.

