---
title: "Sensitivity Analyses for Unmeasured Variables"
author: "Your Name"
output: 
  html_document:
    toc: true
    toc_float: true
---

# Warm-up

Work to understand what the code below is doing. Clarify with the instructor as needed. You will be adapting this code to implement a sensitivity analysis in the context of the smoking-weight gain study that we've explored previously.

```{r}
library(dplyr)
library(ggplot2)

sensitivity_analysis <- function(.data, model_A, model_Y, assoc_A, assoc_Y) {
    n <- nrow(.data)

    # Obtain residuals with residuals()
    # Obtain residual variances with sigma()
    res_A <- residuals(model_A)
    res_var_A <- sigma(model_A)^2
    res_Y <- residuals(model_Y)
    res_var_Y <- sigma(model_Y)^2

    # Compute the mean and variance of U given A and Y
    mean_U_term1 <- (assoc_A/res_var_A)*res_A
    mean_U_term2 <- (((res_var_A - assoc_A^2)*assoc_Y)/(res_var_A*res_var_Y))*res_Y
    mean_U <- mean_U_term1 + mean_U_term2

    var_U_term1 <- (res_var_A - assoc_A^2)/(res_var_A*res_var_Y)
    var_U_term2 <- res_var_Y - assoc_Y^2 + ((assoc_A*assoc_Y)^2)/res_var_A
    var_U <- var_U_term1*var_U_term2

    # Simulate U and add it to the data
    U <- rnorm(n, mean = mean_U, sd = sqrt(var_U))
    .data$U <- U

    ########################################################################
    # The part below is the only part you need to change to implement
    # the sensitivity analysis in a new context.

    # Refit model to estimate the causal effect 
    updated_model <- lm(Y ~ A+C+U, data = .data)
    # The names of the coefficients and confidence interval output rows
    # are called "A" for the treatment variable A.
    # This will change in a new dataset.
    list(c(
        estimate = unname(coefficients(updated_model)["A"]), 
        ci_95_lower = confint(updated_model)["A",1],
        ci_95_upper = confint(updated_model)["A",2]
    ))
}

# Set up simulated example data
set.seed(451)
n <- 10000
U <- rnorm(n, 10, 1)
C <- rnorm(n, 10, 1)
A <- rnorm(n, 2+C+U, 2)
Y <- rnorm(n, 10 + A + C + U, 10)

sim_data <- data.frame(U, C, A, Y)

# Begin the sensitivity analysis

# Fit required models for the sensitivity analysis
mod_A <- lm(A ~ C, data = sim_data)
mod_Y <- lm(Y ~ A+C, data = sim_data)

# Set up degree of association between U and A and between U and Y
# The U->A associations have some constraints: we set up values 
# for the U->A associations that are at most equal to the
# standard deviation of the residuals from the model for A.
U_A_assocs <- seq(from = 0.01, to = sigma(mod_A), length.out = 10)
U_Y_assocs <- seq(from = 0.5, to = 5, by = 0.5)

# Form all combinations of the U->A and U->Y sensitivity parameters
sens_data <- expand.grid(U_A = U_A_assocs, U_Y = U_Y_assocs)

# Run sensitivity analysis
sens_data <- sens_data %>%
    group_by(U_A, U_Y) %>%
    mutate(sens = sensitivity_analysis(sim_data, mod_A, mod_Y, U_A, U_Y))
# Collect sensitivity analysis results in a data.frame
sens_data <- bind_cols(sens_data[,1:2], bind_rows(sens_data$sens))


# Plot results
prepender <- function(string, prefix = "U -> Y strength:") paste(prefix, string)

ggplot(sens_data, aes(x = U_A, y = estimate)) +
    geom_ribbon(aes(ymin = ci_95_lower, ymax = ci_95_upper), fill = "grey70") +
    geom_line() +
    geom_hline(aes(yintercept = coefficients(mod_Y)["A"]), color = "blue", lty = "dashed") +
    geom_hline(aes(yintercept = 0), color = "black") +
    facet_wrap(~U_Y, labeller = as_labeller(prepender)) +
    labs(x = "Strength of U -> A association", y = "ACE and 95% CI")
```

# Main exercise

Here, we will adapt the code above to implement a sensitivity analysis for unmeasured confounding for the smoking-weight gain study.

- The treatment variable is `smkintensity82_71` the change in number of cigarettes smoked per day from 1971 to 1982.
- Assume that a graphical analysis has led to the conclusion that conditional exchangeability holds given `smokeyrs`, `age`, and `asthma`.
- Display graphical output from the sensitivity analysis and write a brief discussion of conclusions you can make from the sensitivity analysis (no more than 400 words).

```{r}
# Load data
library(cidata)
data(nhefs_complete)
# View the data codebook
View(nhefs_codebook)
```
